Vagueness, Noise, and Signaling

### 2015-01-17 paper/version_02/paper.tex
Small changes after first pass. Made introduction into a section. Changed general attitude towards quantal response.

\begin{abstract}
  Signaling games are popular models for studying the evolution of
  meaning, but typical approaches do not incorporate vagueness as a
  feature of successful signaling.  Complementing recent like-minded
  models, we introduce the replicator diffusion dynamic as a special
  case of the replicator mutator dynamic that combines evolutionary
  pressure on optimal signal use with stochastic noise in the
  processing of similar stimuli. Applying this new dynamic to a
  generalization of Lewis' signaling games, we show that stochastic
  imprecision leads to vague, yet by-and-large communicative efficient
  signal use, and, moreover, unifies evolutionary outcomes and helps
  avoid suboptimal categorization.
\end{abstract}

@@

But there are other intriguing aspects about vagueness. The puzzle
that we are concerned with here is how vagueness could arise and be
maintained in the first place. This is a serious worry to
functionalist accounts that maintain that concepts and linguistic
meanings evolved in order to be efficient. Since the existence of
unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward
by \citet{Lipman2009:Why-is-Language}, is to explain how vagueness can
persist under evolutionary pressure to be optimally discriminative. We
address \emph{Lipman's problem} as a technical challenge, namely to
find a conceptually sound and mathematically coherent formal model of
the evolution of categorization and language use that yields
by-and-large efficient categories \emph{and} vagueness.

A number of authors have recently tried to explain why vagueness
evolved as something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of
limitations in information processing or particular learning
strategies
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o}. This
paper contributes to the latter line of thought. Concretely, we
introduce the replicator diffusion dynamics ---a novel variant of the
replicator mutator dynamic--- that integrates stochastic noise on the
differential confusability of similar stimuli. The dynamic proposed
here generalizes and complements previous like-minded accounts. We
show that stochastic noise can not only lead to vague, yet relatively
efficient signal use, but can also unify evolutionary outcomes and
help avoid suboptimal categorization.

@@

An alternative approach is taken by
\citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o} who show how the above desiderata
can be met by evolving strategies in sim-max games. The approach we
take here is similar in spirit, but different in relevant detail. We
are argue here that \citeauthor{FrankeJager2010:Vagueness-Signa}'s
approach builds on a dubious assumption about the source of
vagueness. While \citeauthor{OConnor2013:The-Evolution-o} gives an
agent-level learning dynamic under which vague signaling arises, we
complement the picture by giving a population-level dynamic that
abstracts from the concrete behavior of signalers. A more in-depth
comparison is deferred until Section~\ref{sec:discussion}. Let us
first look into our own proposal in more detail.

@@

The formulation of the \rdd given and explored here is in terms of
behavioral strategies, not in terms of mixed strategies (more on this
below). Dynamics on behavioral strategies assume that agents can adapt
their behavior locally, so to speak, i.e., independently at each
choice point. This greatly reduces the complexity of the dynamic and
simplifies numerical simulations. But it also points towards an
interpretation as a process of cultural evolution, rather than
biological evolution: signaling strategies are not inherited as a
whole, but adapted piece-meal, e.g., by conditional imitation of
``local'' behavior at a given choice point (more below). Still, the
approach we suggest here is compatible with a formulation in terms of
mixed strategies and an interpretation as biological evolution as
well. To show this, we translate our initial formulation of the \rdd
into one using mixed strategies. Doing so identifies the \rdd as a
special case of the \rmd.

@@

Fix a signaling game with finite states $\States$, messages $\Messgs$
and acts $\Acts$. Let $\Pr(\cdot) \in \Delta{\States}$ be the prior
distribution over states and $\Util_{\sen,\rec} \mycolon \States
\times \Messgs \times \Acts \rightarrow \mathds{R}$ the senders's and
receiver's utility functions. A behavioral strategy is a function that
maps an agent's choice points to a probability distribution over
available choices. The sender's behavioral strategies are functions
$\Sstrat \in \Delta(\Messgs)^\States$; the receiver's are functions
$\Rstrat \in \Delta(\Acts)^\Messgs$. The expected utility of choices
at each choice point is:
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot U_\sen(\state, \messg, \act) \\
  \EU(\act, \messg, \Sstrat) & = \sum_{\state \in
    \States} \Pr(\state) \cdot \Sstrat(\messg \probbar \state) \cdot
  U_\rec(\state, \messg, \act) \,.
\end{align*}
The \emph{fitness} of a behavioral strategy at a choice point is the
frequency-weighted average of expected utilities of each choice, given
the opponent's strategy:
\begin{align*}
  \Phi(\state,\Sstrat, \Rstrat) & = \sum_{\messg} \Sstrat(\messg \probbar \state) \cdot
\EU(\messg, \state,\Rstrat) \\
\Phi(\messg,\Sstrat, \Rstrat) & = \sum_{\act} \Rstrat(\act \probbar \messg)
\cdot \EU(\act, \messg,\Sstrat)\,.
\end{align*}
The discrete-time replicator dynamic maps current strategies $\Sstrat$
and $\Rstrat$ to future strategies $\RD(\Sstrat)$ and $\RD(\Rstrat)$
in such a way that changes in frequency are proportional to expected
utilities. For behavioral strategies, the changes take place locally
at each of the agents' choice points:
\begin{align*}
  \RD(\Sstrat)(\messg \probbar \state) & = \frac{\Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)} {\Phi(\state,\Sstrat, \Rstrat)} \\
    \RD(\Rstrat)(\act \probbar \messg) & = \frac{\Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)} {\Phi(\messg,\Sstrat, \Rstrat)}  \,.
\end{align*}
The \rd in behavioral strategies describes the most likely path of
development of average behavior in a (huge and homogeneous) population
of agents when behavioral adaptations occur independently at each
choice point, e.g., when agents adopt the behavior of another agent
\emph{at a given choice point} if that ``local'' behavior is
``locally'' better than their own (see
\citeauthor{Sandholm2013:Population-Game}
\citeyear{Sandholm2013:Population-Game} for agent-level
interpretations of the replicator dynamics and
\citeauthor{Cressman2003:Evolutionary-Dy}
\citeyear{Cressman2003:Evolutionary-Dy} for applications to extensive
form games and behavioral strategies).

@@

The most natural interpretation of this sequential definition is that
the \rdd captures the most likely path of development of aggregate
behavior in a (huge \& homogeneous) society, in which behavioral
changes occur independently at each choice point, as described by the
\rd for behavioral strategies. Whatever strategy an agent does
actually hold, its realization is perturbed by imperfect
discriminability of states in such a way that the expected overall
aggregate behavior is the diffusion of the strategies that the agents
would play in a noise-free world. Any agent-level interpretation of
the \rd in behavioral strategies seems compatible with this
interpretation.

@@

Based on Equations~(\ref{eq:construction-rec}) and
(\ref{eq:construction-sen}), we can formulate the desired
correspondence result. A proof is given in Appendix~\ref{sec:proofs}.

\begin{theorem}
  \label{thm:Correspondence}
  (i) If $G(\Smixed) = \Sstrat$, then $G(\Mutate_{Q^C}(\Smixed)) =
  \Diff_C(\Sstrat)$. And, (ii) if $G(\Rmixed) = \Rstrat$, then
  $G(\Mutate_{R^C}(\Rmixed)) = \Diff_C(\Rstrat)$.
\end{theorem}

@@

\subsection{(Iterated) Diffusion}
\label{sec:iterated-diffusion}

@@

To further explore our simulation results, we track how vague,
communicatively efficient and generally well-structured the recorded
strategy pairs were. \emph{Entropy} captures the amount of
systematicity or regularity in signal use. \emph{Convexity} captures
whether a behavioral strategy would project onto a convex pure
strategy. \emph{Expected utility} measures the communicative
efficiency of evolved strategy pairs.\todo{take average of entropy}

\paragraph{Entropy.} This classic information-theoretic notion
captures the amount of uncertainty in a probability
distribution. Roughly put, entropy of a signaling strategy captures
inverse distance from a pure strategy. The usual definition of entropy
applies directly to mixed strategies, but provably equivalent metrics
for behavioral strategies are ready to hand:
\begin{align*}
  E(\Sstrat) = -\sum_{\state \in \States} \sum_{\messg \in \Messgs}
  \Sstrat(\messg \probbar \state) \cdot \log(\Sstrat(\messg \probbar
  \state)) \\
  E(\Rstrat) = -\sum_{\messg \in \Messgs} \sum_{\state \in \States}
  \Rstrat(\state \probbar \messg) \cdot \log(\Rstrat(\state \probbar
  \messg)) \,. 
\end{align*}
Values obtained by these definitions are lower bounded by $0$ and
upper bounded by, respectively, $\log(|\Messgs^\States|) = |\States|
\cdot \log(|\Messgs|)$ and $\log(|\States^\Messgs|) = |\Messgs| \cdot
\log(|\States|)$. We work with normalized values. The sender
strategies in Figures~\ref{fig:example_stratsA} and
\ref{fig:example_stratsB} have entropy $1.19e^{-5}$ and $0.21$,
respectively. The receiver strategies have respective entropies $0.43$
and $0.84$. In general, we expected that vague languages would have
higher entropy than crisp ones and that increasing imprecision would
increase entropy, all else equal.

\paragraph{Convexity.} At least for sender strategies, which converge
faster, it also makes sense to define a categorical measure of
convexity that compensates for potential vagueness. To determine
whether a sender strategy $\Sstrat$ is convex despite possibly being
vague, we look at the derived pure strategy $\Spure$ for which
$\Spure(\state) = \argmax_{\messg' \in \Messgs}
\Sstrat(\state,\messg')$. If that $\Spure$ is convex, we also count
$\Sstrat$ as convex. The sender strategy in
Figure~\ref{fig:example_stratsA} is not convex, while the one in
Figure~\ref{fig:example_stratsB} is.  One would perhaps not expect a
systematic relation between vagueness and imprecision, on the one
hand, and convexity on the other.

@@

As expected, increasing imprecision leads to higher entropy and lower
expected utility. Importantly, however, imprecision does not
necessarily lead to disastrous decline of communicative success. On
the other hand, imprecision increased the number of outcomes with
convex sender strategies, and led to higher proportions of
convergence. In fact, sufficient impairment always ensured convergence
and convexity. It appears that perceptual imprecision leads to
slightly less communicative efficiency, more vagueness but more
regular, well-behaved languages in shorter time.

Beyond promoting convexity, diffusion also has another interesting
regularizing effect on the evolution of signaling. There was very
little variation in the recorded metrics for evolved strategies, at
least for higher values of impairment. On closer inspection, it turns
out that variability in low-impairment conditions is not only due to
non-convergence or non-convexity. Figure~\ref{fig:MoreExample_strats}
gives two more examples of strategy pairs at stopping time. Both are
obtained for the same triple of parameters, both converged and have
convex sender strategies. However, they are not equally efficient. In
fact, the pair in Figure~\ref{fig:example_stratsC} has a normalized
expected utility of $0.99$ while the one in
Figure~\ref{fig:example_stratsD} only has $0.89$.

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by impairment. To see this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $\Sstrat$ with reversed message indices. The
\emph{within group distances}, i.e., the average distances between all
sender strategies obtained for the same parameter values, are plotted
exemplarily on the left in Figure~\ref{fig:GroupMeasures} for $\beta =
0.1$ and $n_s = 10$. Despite minor quantitative differences, the
picture is the same for all other parameter settings that we tested:
with increasing imprecision, the sender strategies evolving under \rdd
were exactly alike (modulo swapping of messages). This means that
perceptual imprecision unifies evolutionary outcomes and guarantees
sender strategies that are not only convex, but also regular in that
they induce a vague category split exactly in the middle of the unit
interval.

@@

The \rdd is not only compatible with several interpretations of how,
at an agent level, the population-level changes in strategy
proportions come about. It is also compatible with several
interpretations of diffusion. Diffusion can be pictured as a flaw or a
virtue. If we think of perceptual imprecision as a hard limitation on
information processing, vagueness is a natural byproduct of language
evolution. Results from numerical simulations show that such
imprecision leads to vagueness, but does not need to undermine the
prospect of evolving regular and by-and-large efficient signaling
behavior. On the other hand, there appears to be a possibly beneficial
side-effect of state confusability: it speeds up convergence to
regularly shaped and convex categories. This suggests that diffusion
of behavior, as we implemented it here, could also be seen as a form
of stimulus-generalization that is evolutionarily beneficial, because
it accelerates the emergence of uniform category formation. This
latter view is also suggested by a like-minded account to which we
turn next.

@@

\paragraph{Quantal response equilibria.}
\citet{FrankeJager2010:Vagueness-Signa} suggested a number of ways in
which information-processing limitations could lead to vague
strategies. The model that is most clearly related to the present
approach uses the notion of a \emph{quantal response}, also known as a
\emph{soft-max function}
\citep[e.g.][]{Luce1959:Individual-Choi,McFadden1976:Quantal-Choice-,GoereeHolt2008:Quantal-Respons}. A
quantal response function is a paramterized generalization of the
classic best response function. For example, if $U \mycolon \Acts
\rightarrow \mathds{R}$ is the measure of expected utility over
choices $\Acts$ of an agent, then a best response function would have
the agent choose $\act$ only if $U(\act) = \max_{\act' \in \Acts}
U(a)$. A quantal response function rather assumes that agents would
choose $\act$ with a probability proportional to $\expo(\lambda \cdot
U(\act))$, where $\lambda$ is a rationality parameter. If $\lambda
\rightarrow \infty$ we retrieve the behavior of the best response
function, but if it is positive but finite, any choice $\act$ will
receive a positive probability, but acts with higher expected utility
will be more likely. The underlying motivation is the assumption that
there is noise in the computation of expected utilities and/or in
maximization of expected utilities. Consequently, choices with almost
equal expected utilities will be chosen with almost equal probability
(for moderate values of $\lambda$).

@@

But not all quantal response equilibria are equally plausible, and the
ones that are not suggest that it is less natural to think of
vagueness as arising from errors in the computation or maximization of
expected utility than that it arises from confusion of similar
states. To see what the problem is, consider
Figure~\ref{fig:exampleQRE_stratsB}, which is the quantal response
equilibrium for a game with lower tolerance $\toler = 0.05$. Unlike
what evolves under \rdd in this case, sender strategies have vague
boundaries also towards the end of the unit intervals. Technically,
this is because quantal responses equalize message use far away from
the ``prototypical'' interpretation, not just in-between categories,
so to speak. This, in turn, is because quantal responses introduce
noise into the decision making at the level of computing or maximizing
expected utility of choices.

That this is conceptually odd shows even more clearly in a case where
the state space is intuitively unbounded, as for instance for the
property ``tall.'' If the usual interpretation of a ``tall man'' peaks
at around, say, 195cm then when meeting a giant of $n$ meters, senders
would, according to the quantal response approach, be ever more
inclined to describe the giant indifferently as either ``tall'' or
``short'' the larger $n$ gets. This is because, as the distance from
the prototype increases for larger $n$, the difference between the
expected utilities of saying ``short'' or ``tall'' will converge to
zero. 
%Whence that the quantal response approach would predict that
%senders would grow indifferent between choice of antonyms as $n$
%grows, which seems weird.

Admittedly, this argument hinges on the choice of utility
function. Still, to the extent that the utility functions we chose are
reasonable---and we think they are very reasonable--- the case
suggests that quantal responses are not a good model, intuitively
speaking, for why linguistic categories are vague. Vagueness is more
plausibly an effect of perceptual confusion of similar states than of
computational errors in maximizing expected utility.

>>>>

\begin{abstract}
  Signaling games are popular models for studying the evolution of
  meaning, but typical approaches do not incorporate vagueness as a
  feature of successful signaling.  Complementing recent like-minded
  models, we introduce the replicator diffusion dynamic as a special
  case of the replicator mutator dynamic that combines evolutionary
  pressure on optimal signal use with stochastic noise in the
  processing of similar stimuli. Applying this new dynamic to a
  generalization of Lewis' signaling games, we show that stochastic
  imprecision leads to vague, yet by-and-large communicative efficient
  signal use, and, moreover, that it unifies evolutionary outcomes and helps
  avoiding suboptimal categorization.
\end{abstract}

\section{Introduction}
\label{sec:introduction}

@@

But there are other intriguing aspects about vagueness. The puzzle
that we are concerned with here is how vagueness could arise and be
maintained in the first place. This is a serious issue for
functionalist accounts that maintain that concepts and linguistic
meanings evolved driven towards efficiency. Since the existence of
unclear borderline cases seems to entail inefficiency of
categorization or communication, the challenge, succinctly put forward
by \citet{Lipman2009:Why-is-Language}, is to explain how vagueness can
persist under evolutionary pressure to be optimally discriminative. We
address \emph{Lipman's problem} as a technical challenge, namely to
find a conceptually sound and mathematically coherent formal model of
the evolution of categorization and language use that yields
by-and-large efficient categorization \emph{and} vagueness.

A number of authors have recently tried to explain why vagueness
evolved as something that is itself useful
\citep[e.g.][]{Jaegherde-Jaegher2003:A-Game-Theoreti,Deemter2009:Utility-and-Lan,BlumeBoard2013:Intentional-Vag}.
Others have argued that vagueness is a natural byproduct of
limitations in information processing or particular learning
strategies
\citep[e.g.][]{FrankeJager2010:Vagueness-Signa,OConnor2013:The-Evolution-o}. This
paper contributes to the latter line of thought. Concretely, we
introduce the replicator diffusion dynamics---a novel variant of the
replicator mutator dynamic---that integrates stochastic noise on the
differential confusability of similar stimuli. The dynamic proposed
here generalizes and complements previous like-minded accounts. We
show that stochastic noise can not only lead to vague, yet relatively
efficient signal use, but also unify evolutionary outcomes and
help avoid suboptimal categorization.

@@

An alternative approach is taken by
\citet{FrankeJager2010:Vagueness-Signa} and
\citet{OConnor2013:The-Evolution-o} who show how the above desiderata
can be met by evolving strategies in sim-max games. The approach we
take here is similar in spirit, but different in relevant detail. We
argue here that \citeauthor{FrankeJager2010:Vagueness-Signa}'s
approach gives rise to problematic predictions under certain conditions
which are avoided in our approach. However, the two are not necessarily
competing explanations since they are motivated differently and it is
certainly possible that both the mechanisms proposed by them, as the
ones proposed here play a role in the whole phenomenon.
Regarding the work by \citeauthor{OConnor2013:The-Evolution-o}, while she
gives an agent-level learning dynamic under which vague signaling arises,
we complement the picture by giving a population-level dynamic that
abstracts from the concrete behavior of signalers. A more in-depth
comparison is deferred until Section~\ref{sec:discussion}. Let us
first look into our own proposal in more detail.

@@

The formulation of the \rdd given and explored here is in terms of
behavioral strategies, not in terms of mixed strategies (more on this
below). Dynamics on behavioral strategies assume that agents can adapt
their behavior locally, so to speak, i.e., independently at each
choice point. This greatly reduces the complexity of the dynamic and
simplifies numerical simulations. But it also points towards an
interpretation of the process as cultural evolution, rather than
biological evolution. Signaling strategies are not inherited as a
whole, but adapted piece-meal, e.g., by conditional imitation of
``local'' behavior at a given choice point (more below). Still, the
approach we suggest here is compatible with a formulation in terms of
mixed strategies and an interpretation as biological evolution as
well. To show this, we translate our initial formulation of the \rdd
into one using mixed strategies. Doing so identifies the \rdd as a
special case of the \rmd.

@@

Fix a signaling game with finite states $\States$, messages $\Messgs$
and acts $\Acts$. Let $\Pr(\cdot) \in \Delta{\States}$ be the prior
distribution over states and $\Util_{\sen,\rec} \mycolon \States
\times \Messgs \times \Acts \rightarrow \mathds{R}$ the senders's and
receiver's utility functions. A behavioral strategy is a function that
maps an agent's choice points to a probability distribution over
available choices. The sender's behavioral strategies are functions
$\Sstrat \in \Delta(\Messgs)^\States$; the receiver's are functions
$\Rstrat \in \Delta(\Acts)^\Messgs$. The expected utility of choices
at each choice point is:
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot U_\sen(\state, \messg, \act) \\
  \EU(\act, \messg, \Sstrat) & = \sum_{\state \in
    \States} \Pr(\state) \cdot \Sstrat(\messg \probbar \state) \cdot
  U_\rec(\state, \messg, \act) \,.
\end{align*}
The \emph{fitness} of a behavioral strategy at a choice point is the
frequency-weighted average of expected utilities of each choice, given
the opponent's strategy:
\begin{align*}
  \Phi(\state,\Sstrat, \Rstrat) & = \sum_{\messg} \Sstrat(\messg \probbar \state) \cdot
\EU(\messg, \state,\Rstrat) \\
\Phi(\messg,\Sstrat, \Rstrat) & = \sum_{\act} \Rstrat(\act \probbar \messg)
\cdot \EU(\act, \messg,\Sstrat)\,.
\end{align*}
The discrete-time replicator dynamic maps current strategies $\Sstrat$
and $\Rstrat$ to future strategies $\RD(\Sstrat)$ and $\RD(\Rstrat)$
in such a way that changes in frequency are proportional to expected
utilities. For behavioral strategies, the changes take place locally
at each of the agents' choice points:
\begin{align*}
  \RD(\Sstrat)(\messg \probbar \state) & = \frac{\Sstrat(\messg \probbar \state) \cdot
    \EU(\messg, \state,\Rstrat)} {\Phi(\state,\Sstrat, \Rstrat)} \\
    \RD(\Rstrat)(\act \probbar \messg) & = \frac{\Rstrat(\act \probbar \messg) \cdot
    \EU(\act, \messg,\Sstrat)} {\Phi(\messg,\Sstrat, \Rstrat)}  \,.
\end{align*}
The \rd in behavioral strategies describes the most likely path of
development of average behavior in a population
of agents when behavioral adaptations occur independently at each
choice point, e.g., when agents adopt the behavior of another agent
\emph{at a given choice point} if that ``local'' behavior is
``locally'' better than their own (see
\citeauthor{Sandholm2013:Population-Game}
\citeyear{Sandholm2013:Population-Game} for agent-level
interpretations of the replicator dynamics and
\citeauthor{Cressman2003:Evolutionary-Dy}
\citeyear{Cressman2003:Evolutionary-Dy} for applications to extensive
form games and behavioral strategies).

@@

The most natural interpretation of this sequential definition is that
the \rdd captures the most likely path of development of aggregate
behavior in a society, in which behavioral
changes occur independently at each choice point, as described by the
\rd for behavioral strategies. Whatever strategy an agent does
actually hold, its realization is perturbed by imperfect
discriminability of states in such a way that the expected overall
aggregate behavior is the diffusion of the strategies that the agents
would play in a noise-free world. Any agent-level interpretation of
the \rd in behavioral strategies seems compatible with this
interpretation.

@@

Based on Equations~(\ref{eq:construction-rec}) and
(\ref{eq:construction-sen}), we can formulate the desired
correspondence result, a proof for which is given in Appendix~\ref{sec:proofs}.

\begin{theorem}
  \label{thm:Correspondence}
  (i) If $G(\Smixed) = \Sstrat$, then $G(\Mutate_{Q^C}(\Smixed)) =
  \Diff_C(\Sstrat)$. (ii) If $G(\Rmixed) = \Rstrat$, then
  $G(\Mutate_{R^C}(\Rmixed)) = \Diff_C(\Rstrat)$.
\end{theorem}

@@

\subsection{Iterated diffusion}
\label{sec:iterated-diffusion}

@@

To further explore our simulation results, we calculated
metrics that aim to numerically capture how vague, generally
well-structured, and communicatively efficient the recorded
strategy pairs were. \emph{Entropy} captures the amount of
systematicity or regularity in signal use. \emph{Convexity} captures
whether a behavioral strategy would project onto a convex pure
strategy. \emph{Expected utility} measures the communicative
efficiency of evolved strategy pairs.\todo{take average of entropy}

\paragraph{Entropy.} This classic information-theoretic notion
captures the amount of uncertainty in a probability
distribution. Roughly put, entropy of a signaling strategy captures
inverse distance from a pure strategy. The usual definition of entropy
applies directly to mixed strategies, but provably equivalent metrics
for behavioral strategies are ready to hand:
\begin{align*}
  E(\Sstrat) = -\sum_{\state \in \States} \sum_{\messg \in \Messgs}
  \Sstrat(\messg \probbar \state) \cdot \log(\Sstrat(\messg \probbar
  \state)) \\
  E(\Rstrat) = -\sum_{\messg \in \Messgs} \sum_{\state \in \States}
  \Rstrat(\state \probbar \messg) \cdot \log(\Rstrat(\state \probbar
  \messg)) \,. 
\end{align*}
Values obtained by these definitions are lower bounded by $0$ and
upper bounded by, respectively, $\log(|\Messgs^\States|) = |\States|
\cdot \log(|\Messgs|)$ and $\log(|\States^\Messgs|) = |\Messgs| \cdot
\log(|\States|)$. We work with normalized values. The sender
strategies in Figures~\ref{fig:example_stratsA} and
\ref{fig:example_stratsB} have entropy $1.19e^{-5}$ and $0.21$,
respectively. The receiver strategies have respective entropies $0.43$
and $0.84$. In general, we expected that vague languages would have
higher entropy than crisp ones and that increasing imprecision would
lead to increased entropy, all else being equal.

\paragraph{Convexity.} At least for sender strategies, which converge
faster, it also makes sense to define a categorical measure of
convexity that compensates for potential vagueness. To determine
whether a sender strategy $\Sstrat$ is convex despite possibly being
vague, we look at the derived pure strategy $\Spure$ for which
$\Spure(\state) = \argmax_{\messg' \in \Messgs}
\Sstrat(\state,\messg')$. If that $\Spure$ is convex, we also count
$\Sstrat$ as convex. The sender strategy in
Figure~\ref{fig:example_stratsA} is not convex, while the one in
Figure~\ref{fig:example_stratsB} is.  One would a priori not expect a
systematic relation between imprecision and convexity.

@@

As expected, increasing imprecision leads to higher entropy and lower
expected utility. Importantly, however, imprecision does not
necessarily lead to disastrous decline of communicative success. What was
not expected, on the other hand, was a relation between imprecision and
convexity. However, the corresponding plot clearly shows that
higher imprecision led to a higher number of outcomes with
convex sender strategies. Moreover, it also led to higher likelihood of
convergence. In fact, sufficient impairment always ensured convergence
and convexity. It appears that perceptual imprecision leads to
more entropy, slightly less communicative efficiency, but more
regular, well-behaved languages in shorter time.

Beyond promoting convexity and convergence, diffusion also has another interesting
regularizing effect on the evolution of signaling. There was very
little variation in the recorded metrics for evolved strategies, at
least for higher values of impairment. On closer inspection, it turns
out that variability in low-impairment conditions is not only due to
non-convergence or non-convexity. Figure~\ref{fig:MoreExample_strats}
gives two more examples of strategy pairs at stopping time. Both are
obtained for the same triple of parameters, both converged before the
maximum number of rounds, and both have
convex sender strategies. However, they are not equally efficient. In
fact, the pair in Figure~\ref{fig:example_stratsC} has a normalized
expected utility of $0.99$ while the one in
Figure~\ref{fig:example_stratsD} only has $0.89$.

@@

Interestingly, this type of variability in evolutionary outcomes can
be weeded out by impairment. To investigate this, we calculated the average
distance between evolved sender strategies within each group of trials
that had identical parameter values. We determined the distance
between strategies $\Sstrat$ and $\Sstrat'$ as the average Hellinger
distance between distributions $\Sstrat(\state)$ and
$\Sstrat'(\state)$ at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $\Sstrat$ with reversed message indices. An example of the
\emph{within group distances}, i.e., the average distances between all
sender strategies obtained for the same parameter values, is plotted
on the left in Figure~\ref{fig:GroupMeasures} for $\beta =
0.1$ and $n_s = 10$. Despite some quantitative differences, the
general trend is the same for all other parameter settings that we tested:
with increasing imprecision, the sender strategies evolving under \rdd
were much more alike (modulo swapping of messages). This means that
perceptual imprecision unifies evolutionary outcomes and guarantees
sender strategies that are not only convex, but also regular in that
they induce a vague category split exactly in the middle of the unit
interval.

@@

The \rdd is not only compatible with several interpretations of how,
at an agent level, the population-level changes in strategy
proportions come about. It is also compatible with several
interpretations of diffusion. Diffusion can be pictured as a flaw or a
virtue. If we think of perceptual imprecision as a hard limitation on
information processing, vagueness is a natural byproduct of language
evolution. Results from numerical simulations show that such
imprecision leads to vagueness, but does not need to undermine the
prospect of evolving regular and by-and-large efficient signaling
behavior. On the other hand, there appears to be a possibly beneficial
side-effect of state confusability: it speeds up convergence to
regularly shaped and convex categories. This suggests that diffusion
of behavior, as we implemented here, could also be seen as a form
of stimulus-generalization that is evolutionarily beneficial, because
it accelerates the emergence of uniform category formation. This
latter view is also suggested by a like-minded account to which we
turn next.

@@

\paragraph{Quantal response equilibria.}
\citet{FrankeJager2010:Vagueness-Signa} suggested a number of ways in
which information-processing limitations could lead to vague
strategies. The model that is most clearly related to the present
approach uses the notion of a \emph{quantal response}, also known as a
\emph{soft-max function}
\citep[e.g.][]{Luce1959:Individual-Choi,McFadden1976:Quantal-Choice-,GoereeHolt2008:Quantal-Respons}. A
quantal response function is a paramterized generalization of the
classic best response function. For example, if $U \mycolon \Acts
\rightarrow \mathds{R}$ is the measure of expected utility over
choices $\Acts$ of an agent, then a best response function would have
the agent choose $\act$ only if $U(\act) = \max_{\act^\prime \in \Acts}
U(\act^\prime)$. A quantal response function rather assumes that agents would
choose $\act$ with a probability proportional to $\expo(\lambda \cdot
U(\act))$, where $\lambda$ is a rationality parameter. If $\lambda
\rightarrow \infty$ we retrieve the behavior of the best response
function, but if it is positive but finite, any choice $\act$ will
receive a positive probability, but acts with higher expected utility
will be more likely. The underlying motivation is the assumption that
there is noise in the computation of expected utilities and/or in
maximization of expected utilities. Consequently, choices with almost
equal expected utilities will be chosen with almost equal probability
(for moderate values of $\lambda$).

@@

But not all quantal response equilibria are equally plausible.
%, and the
%ones that are not suggest that it is less natural to think of
%vagueness as arising from errors in the computation or maximization of
%expected utility than that it arises from confusion of similar states
To see what the problem is, consider
Figure~\ref{fig:exampleQRE_stratsB}, which is the quantal response
equilibrium for a game with lower tolerance $\toler = 0.05$. Unlike
what evolves under \rdd in this case, sender strategies have vague
boundaries also towards the end of the unit intervals. Technically,
this is because quantal responses equalize message use far away from
the ``prototypical'' interpretation, not just in-between categories,
so to speak. This, in turn, is because quantal responses introduce
noise into the decision making at the level of computing or maximizing
expected utility of choices.

That this is problematic shows even more clearly in a case where
the state space is intuitively unbounded, as for instance for the
property ``tall.'' If the usual interpretation of a ``tall man'' peaks
at around, say, 195cm then when meeting a giant of $n$ meters, senders
would, according to the quantal response approach, be ever more
inclined to describe the giant indifferently as either ``tall'' or
``short'' the larger $n$ gets. This is because, as the distance from
the prototype increases for larger $n$, the difference between the
expected utilities of saying ``short'' or ``tall'' will converge to
zero. 
%Whence that the quantal response approach would predict that
%senders would grow indifferent between choice of antonyms as $n$
%grows, which seems weird.

Admittedly, this argument hinges on the choice of utility
function. Still, to the extent that the utility functions we chose are
reasonable---and we think they are very reasonable---the case
suggests that quantal responses do not paint a complete picture. Naturally,
we cannot claim our approach provides valid predictions for all cases.
Given the complexity of the phenomenon, and the fact that
the quantal response approach is built on very reasonable assumptions,
it is likely that both proposals are relevant pieces of the puzzle.
%Vagueness is more
%plausibly an effect of perceptual confusion of similar states than of
%computational errors in maximizing expected utility.
